Transitivity of Subtyping for Intersection Types

### 2019-06-23 papers/intersection-transitivity.tex
check

\begin{figure}[tbp]
  \fbox{$A ≤ B$}
  \begin{gather*}
    \text{(refl)} \; \inference{}{A ≤ A} \quad
    \text{(trans)} \; \inference{A ≤ B & B ≤ C}{A ≤ C} \\[3ex]
    \text{(incl$_L$)} \; \inference{}{A ∩ B ≤ A} \quad
    \text{(incl$_R$)} \; \inference{}{A ∩ B ≤ B} \quad
    \text{(glb)} \; \inference{C ≤ A & C ≤ B}{C ≤ A ∩ B} \\[3ex]
    (→) \; \inference{A' ≤ A & B ≤ B'}{A → B ≤ A' → B'} \quad
    ({→}{∩}) \; \inference{}{(A → B) ∩ (A → C) ≤ A → (B ∩ C)} \\[3ex]
    (\TOP_{\mathrm{top}}) \; \inference{}{A ≤ \TOP} \quad
    (\TOP{→}) \; \inference{}{\TOP ≤ A → \TOP}
  \end{gather*}
  \caption{Subtyping of Barendregt, Coppo, and
    Dezani-Ciancaglini (BCD).}
  \label{fig:BCD-subtyping}
\end{figure}

@@

Our new subtyping judgment relies on several auxilliary notions that
help us avoid the use of ellipses, which we define in
Figure~\ref{fig:aux}. These include the $\dom{A}$ and $\cod{A}$
functions, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
relations $A ∈ B$ and $A ⊆ B$.
%
The $\dom{A}$ and $\cod{A}$ functions return the domain or codomain if
$A$ is a function type, respectively. If $A$ is an intersection $A_1 ∩
A_2$, then $\dom{A}$ is the intersection of the domain of $A_1$ and
$A_2$.  If $A$ is an atom, $\dom{A}$ is undefined. Likewise for
$\cod{A}$. For example, if $A = (A_1 → B_1) ∩ \cdots ∩ (A_n → B_n)$,
then $\dom{A} = A_1 ∩ \cdots ∩ A_n$ and $\cod{A} = B_1 ∩ \cdots ∩ B_n$.
%
The $\topP{A}$ predicate identifies types that are equivalent to
$\TOP$. The $\topInCod{A}$ predicate identifies types that have $\TOP$
in their codomain.
%
The relation $A ∈ B$ indicates whether $A$ is syntactically a part of $B$.
The relation $A ⊆ B$ holds when every part of $A$ is a part of $B$.
We say that $B$ \emph{contains} $A$ if $A ⊆ B$.

@@

\begin{figure}[tbp]
  \fbox{$A <: B$}
  \begin{gather*}
    \text{(refl$_α$)} \; \inference{}{α <: α} \\[3ex]
    \text{(lb$_L$)} \; \inference{A <: C}{A ∩ B <: C} \quad
    \text{(lb$_R$)} \; \inference{B <: C}{A ∩ B <: C} \quad
    \text{(glb)} \; \inference{C ≤ A & C ≤ B}{C ≤ A ∩ B} \\[3ex]
    (→') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C → D}
    \begin{array}{l} \neg\, \mathsf{top}(D) \\
      \neg \, \mathsf{topInCod}(B) \\
      B ⊆ A \end{array}\\[3ex]
    (\TOP_{\mathrm{top}}) \; \inference{}{A <: \TOP} \quad
    (\TOP{→}') \; \inference{}{C <: A → B}\;\mathsf{top}(B)
  \end{gather*}
  \caption{The New Subtyping Judgment}
  \label{fig:new-subtyping}
\end{figure}

The new intersection subtyping judgment, $A <: B$, is defined in
Figure~\ref{fig:new-subtyping}. Of course, it omits the (trans) rule.
We also omit the (refl) rule, replacing it with reflexivity for atoms
(refl$_\alpha$). The most important rule is the one for function types
$(→')$, which subsumes the rules $(→)$ and $({→}{∩})$ in BCD
subtyping. It essentially turns the $\beta$-soundness property into a
subtyping rule. It says that a type $A$ is a subtype of a function
type $C → D$ if $A$ contains a type $B$ whose domain and codomain are
larger and smaller than $C$ and $D$, respectively. The use of $B$
instead of $A$ in the premise, with the condition $B ⊆ A$, allows this
rule to absorb uses of (incl$_L$) and (incl$_R$) on the left.  The
side conditions $\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed
because of the (\TOP{→}') rule, which in turn is needed to preserve
types under $\eta$-reduction.  In a system that does not involve
$\eta$-reduction, the (\TOP{→}') rule can be ommitted, as well as
those side conditions. The rules (lb$_L$) and (lb$_R$) adapt
(incl$_L$) and (incl$_R$) to a system without transitivity, and have
appeared many times in the literature~\citep{Bakel:1995aa}.

\begin{proposition}\ 
  \begin{enumerate}
  \item (reflexivity) $A <: A$ \label{prop:refl}
  \item If $A <: B ∩ C$, then $A <: B$ and $A <: C$. \label{prop:⊔⊑-inv}
  \end{enumerate}
\end{proposition}
\begin{proof}
  The proof of reflexivity is by induction on $A$. In the case
  $A = A_1 → A_2$, we proceed by cases on whether $\topP{A_2}$.
  If it is, deduce $A_1 → A_2 <: A_1 → A_2$ by rule $(\TOP{→}')$.
  Otherwise, we use rule $(→')$
\end{proof}

\begin{proposition}\label{prop:u⊆v⊑w→u⊑w}
 If $A <: B$ and $C ⊆ B$, then $A <: C$.
\end{proposition}

\clearpage
\pagebreak
\section{Transitivity}

\begin{definition}[factors]
  We say $A → B$ \emph{factors} $C$
  if there exists some type $C'$ such that
  $C' ⊆ C$, $\neg\,\mathsf{topInCod}(C')$, $A <: \dom{C'}$, and $\cod{C'} <: B$.
\end{definition}

\begin{lemma}\label{lem:⊑-fun-inv}
  If $\neg\,\mathsf{top}(B)$,
  $D <: C$, and
  $A → B ∈ C$, then
  $A → B$ factors $D$.
\end{lemma}

\begin{lemma}\label{lem:sub-inv-trans}
  If
  \begin{itemize}
  \item $\neg\, \mathsf{topInCod}(D)$,
  \item for any $A,B$, if $\neg\,\mathsf{top}(B)$ and $A → B ∈ D$,
    then $A → B$ factors $C$,
  \end{itemize}
  then $\dom{D} → \cod{D}$ factors $C$.
\end{lemma}

>>>>

\begin{figure}[tbp]
  \fbox{$A ≤ B$}
  \begin{gather*}
    \text{(refl)} \; \inference{}{A ≤ A} \quad
    \text{(trans)} \; \inference{A ≤ B & B ≤ C}{A ≤ C} \\[3ex]
    \text{(incl$_L$)} \; \inference{}{A ∩ B ≤ A} \quad
    \text{(incl$_R$)} \; \inference{}{A ∩ B ≤ B} \quad
    \text{(glb)} \; \inference{A ≤ C & A ≤ D}{A ≤ C ∩ D} \\[3ex]
    (→) \; \inference{C ≤ A & B ≤ D}{A → B ≤ C → D} \quad
    ({→}{∩}) \; \inference{}{(A → B) ∩ (A → C) ≤ A → (B ∩ C)} \\[3ex]
    (\TOP_{\mathrm{top}}) \; \inference{}{A ≤ \TOP} \quad
    (\TOP{→}) \; \inference{}{\TOP ≤ C → \TOP}
  \end{gather*}
  \caption{Subtyping of Barendregt, Coppo, and
    Dezani-Ciancaglini (BCD).}
  \label{fig:BCD-subtyping}
\end{figure}

@@

Our new subtyping judgment relies on several auxilliary notions that
help us avoid the use of ellipses, which we define in
Figure~\ref{fig:aux}. These include the $\dom{A}$ and $\cod{A}$
functions, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
relations $A ∈ B$ and $A ⊆ B$.
%
The $\dom{A}$ and $\cod{A}$ functions return the domain or codomain if
$A$ is a function type, respectively. If $A$ is an intersection $A_1 ∩
A_2$, then $\dom{A}$ is the intersection of the domain of $A_1$ and
$A_2$.  If $A$ is an atom, $\dom{A}$ is undefined. Likewise for
$\cod{A}$. For example, if $A = (A_1 → B_1) ∩ \cdots ∩ (A_n → B_n)$,
then $\dom{A} = A_1 ∩ \cdots ∩ A_n$ and $\cod{A} = B_1 ∩ \cdots ∩
B_n$.  When $\dom{A}$ or $\cod{A}$ appears in lemma or theorem
statement, we implicitly assume that $A$ is a type such that $\dom{A}$
and $\cod{A}$ are defined.
%
The $\topP{A}$ predicate identifies types that are equivalent to
$\TOP$. The $\topInCod{A}$ predicate identifies types that have $\TOP$
in their codomain.
%
The relation $A ∈ B$ indicates whether $A$ is syntactically a part of $B$.
The relation $A ⊆ B$ holds when every part of $A$ is a part of $B$.
We say that $B$ \emph{contains} $A$ if $A ⊆ B$.

@@

\begin{figure}[tbp]
  \fbox{$A <: B$}
  \begin{gather*}
    \text{(refl$_α$)} \; \inference{}{α <: α} \\[3ex]
    \text{(lb$_L$)} \; \inference{A <: C}{A ∩ B <: C} \quad
    \text{(lb$_R$)} \; \inference{B <: C}{A ∩ B <: C} \quad
    \text{(glb)} \; \inference{A ≤ C & A ≤ D}{A ≤ C ∩ D} \\[3ex]
    (→') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C → D}
    \begin{array}{l} \neg\, \mathsf{top}(D) \\
      \neg \, \mathsf{topInCod}(B) \\
      B ⊆ A \end{array}\\[3ex]
    (\TOP_{\mathrm{top}}) \; \inference{}{A <: \TOP} \quad
    (\TOP{→}') \; \inference{}{A <: C → D}\;\mathsf{top}(D)
  \end{gather*}
  \caption{The New Subtyping Judgment}
  \label{fig:new-subtyping}
\end{figure}

The new intersection subtyping judgment, $A <: B$, is defined in
Figure~\ref{fig:new-subtyping}. First, it does not include the (trans)
rule.  It also replaces the (refl) rule with reflexivity for atoms
(refl$_\alpha$). The most important rule is the one for function types
$(→')$, which subsumes $(→)$ and $({→}{∩})$ in BCD subtyping.  The
$(→')$ rule essentially turns the $\beta$-soundness property into a
subtyping rule. It says that a type $A$ is a subtype of a function
type $C → D$ if $A$ contains a type $B$ whose domain and codomain are
larger and smaller than $C$ and $D$, respectively. The use of $B$
instead of $A$ in the premise, with the condition $B ⊆ A$, allows this
rule to absorb uses of (incl$_L$) and (incl$_R$) on the left.  The
side conditions $\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed
because of the (\TOP{→}') rule, which in turn is needed to preserve
types under $\eta$-reduction.  In a system that does not involve
$\eta$-reduction, the (\TOP{→}') rule can be ommitted, as well as
those side conditions. The rules (lb$_L$) and (lb$_R$) adapt
(incl$_L$) and (incl$_R$) to a system without transitivity, and have
appeared many times in the literature~\citep{Bakel:1995aa}.  The
$(\TOP{→}')$ rule generalizes the (\TOP{→}) rule, replacing the $\TOP$
on the left with any type $A$, because for transitivity, any type is
below $\TOP$. The $(\TOP{→}')$ rule also replaces the $\TOP$ in the
codomain on the right with any type $D$ that is equivalent to $\TOP$.

\begin{proposition}\ 
  \begin{enumerate}
  \item (reflexivity) $A <: A$ \label{prop:refl}
  \item If $A <: B ∩ C$, then $A <: B$ and $A <: C$. \label{prop:⊔⊑-inv}
  \item If $A <: B$ and $C ∈ B$, then $A <: C$.\label{prop:u∈v⊑w→u⊑w}
  \item If $A <: B$ and $C ⊆ B$, then $A <: C$.\label{prop:u⊆v⊑w→u⊑w}
  \end{enumerate}
\end{proposition}
\begin{proof}\ 
  \begin{enumerate}
    \item The proof of reflexivity is by induction on $A$. In the case
      $A = A_1 → A_2$, we proceed by cases on whether $\topP{A_2}$.
      If it is, deduce $A_1 → A_2 <: A_1 → A_2$ by rule $(\TOP{→}')$.
      Otherwise, we use rule $(→')$
    \item The proof is by induction on the derivation of $A <: B ∩ C$.
    \item The proof is by induction on $B$. In the case where $B = B_1 ∩ B_2$,
      we have that either $C ∈ B_1$ or $C ∈ B_2$, but in
      either case we use part \ref{prop:⊔⊑-inv} of this proposition to
      fulfill the premise of the induction hypothesis.
    \item The proof  is by induction on $C$, using part \ref{prop:u∈v⊑w→u⊑w}
      of this proposition in the cases for atoms and function types. 
  \end{enumerate}
\end{proof}

Next we turn to the subtyping inversion principle for function types.
The idea is to generalize the rule $(→')$ with respect to the type on
the right, allowing any type that contains a function type.  The
premises of $(→')$ are somewhat complex, so we package most of them
into the following definition.

\begin{definition}[factors]
  We say $C → D$ \emph{factors} $A$
  if there exists some type $B$ such that
  $B ⊆ A$, $\neg\,\mathsf{topInCod}(B)$, $C <: \dom{B}$, and $\cod{B} <: D$.
\end{definition}

\begin{proposition}[Inversion Principle for Function Types]\label{lem:⊑-fun-inv}
  If $A <: C'$, $C → D ∈ C'$, and $\neg\,\mathsf{top}(D)$, then
  $C → D$ factors $A$.
\end{proposition}
\begin{proof}
  The proof is a straightforward induction on $A <: C'$.
\end{proof}

\section{Transitivity}

The proof of transitivity relies on the following lemma, which is
traditionally needed to prove the inversion principle for function
types. However, it was not needed for our system because the rule
$(→')$ is already quite close to the inversion principle.
The lemma states that if every function type $C → D$ in A
factors $B$, then $\dom{A} → \cod{A}$ also factors $B$.

\begin{lemma}\label{lem:sub-inv-trans}
  If
  \begin{itemize}
  \item for any $C$ and $D$, if $C → D ∈ A$ and $\neg\,\mathsf{top}(D)$,
    then $C → D$ factors $B$, and
  \item $\neg\, \mathsf{topInCod}(A)$,
  \end{itemize}
  then $\dom{A} → \cod{A}$ factors $B$.
\end{lemma}
\begin{proof}
  The proof is by induction on $A$.
  \begin{itemize}
  \item Case $A$ is an atom. The statement is vacuously true.
  \item Case $A = A_1 → A_2$ is a function type. Then we conclude by applying
    the premise with $C$ and $D$ instantiated to $A_1$ and $A_2$ respectively.
  \item Case $A = A_1 ∩ A_2$.

    UNDER CONSTRUCTION
  \end{itemize}
\end{proof}
